---
title: "BEscreen_240201"
author: "Andrea"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Nicole used the same guides as Dane's P300 library (ABE & BE4) so reprocessing
# Dane's original screen to compare hits between them
# looking at changes in MYC expression, sorted by high/low neon green -> compare
# each of these to respective controls
# 2x timepoints, 2x replicates, both WT & KO background

fastq dir: /pipeline/Archives/NextSeq/240201_VH01624_66_AAF322FM5/ProjectFolders/Project_Dane-Vassiliadis/Sample*
project dir: /dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis

# align base editor library to guide library fasta 
# aligning with 0 mismatches to start 
```{bash}
mkdir fastqc
mkdir logs
mkdir trimmed_fastqs
mkdir alignment
mkdir session-info
mkdir mapped_guides
mkdir guide_counts

# align to guide library
for fq in /pipeline/Archives/NextSeq/240201_VH01624_66_AAF322FM5/ProjectFolders/Project_Dane-Vassiliadis/Sample*/*fastq.gz
do
sbatch scripts/MapCRISPR.sbatch $fq
done

module load py-multiqc/1.15-gcc-13.2.0
multiqc .
```

# load libraries and set paths
```{r}
library(data.table)
library(limma)

project.dir <- "/dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis"
counts.dir <- file.path(project.dir, "guide_counts")
```

# combine guide counts into a counts table
```{r}
gc.files <- list.files(counts.dir)
all.counts <- fread(file.path(counts.dir, gc.files[1]), skip = 1)
colnames(all.counts) <- c(gc.files[1], "sgRNA")
for(gc in gc.files[-1]){
  counts <- fread(file.path(counts.dir, gc), skip = 1)
  colnames(counts) <- c(gc, "sgRNA")
  all.counts <- merge(all.counts, counts, by = "sgRNA", all = TRUE)
}
colnames(all.counts) <- gsub(".guideCounts.txt", "", colnames(all.counts))

# convert to mageck format
sg.split <- strsplit2(all.counts$sgRNA, "_")
all.counts$gene <- sg.split[, 1]
all.counts <- all.counts[, c(1, 40, 2:39)]

# clear up any issues or formatting that are problematic for mageck
# are there any NAs?
sum(is.na(all.counts))
# [1] 130
# change any NAs to 0 counts 
all.counts[is.na(all.counts)] <- 0
# write out all.counts table
write.table(
  all.counts,
  file.path(counts.dir, "all_guideCounts.txt"),
  sep = "\t",
  col.names = TRUE,
  row.names = FALSE,
  quote = FALSE
)

# filter out very low counts, a couple of problematic guides should be removed
# require at least 100 counts in at least half of the samples
keep <- which(rowSums(all.counts >= 100) >= 19)
# have a look at those guides which failed filter
all.counts[-keep, ]
# these 5 guides are all consistently low across the board

filt.counts <- all.counts[keep, ]
write.table(
  filt.counts,
  file.path(counts.dir, "filtered_guideCounts.txt"),
  sep = "\t",
  col.names = TRUE,
  row.names = FALSE,
  quote = FALSE
)
```


# run mageck test for desired comparisons
```{bash}
# run interactively in Centos container 
sruncentos # locally aliased
source /etc/profile.d/modules.sh

mkdir mageck_counts

module purge
module load mageck/0.5.9
module load scipy/1.7.3
module load R/4.4.2

mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 3-ABE-B2-dP-HI-48h_S20,3-ABE-F1-dP-HI-48h_S57 \
  -c 3-ABE-B2-dP-LIB-48h_S41,3-ABE-F1-dP-LIB-48h_S30 \
  -n mageck_counts/3_ABE_dP_48h_HI_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 3-ABE-B2-dP-LO-48h_S28,3-ABE-F1-dP-LO-48h_S27 \
  -c 3-ABE-B2-dP-LIB-48h_S41,3-ABE-F1-dP-LIB-48h_S30 \
  -n mageck_counts/3_ABE_dP_48h_LO_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
  
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 3-BE4-75-dP-HI-48h_S21,3-BE4-78-dP-HI-48h_S22 \
  -c 3-BE4-75-dP-LIB-48h_S51,3-BE4-78-dP-LIB-48h_S52 \
  -n mageck_counts/3_BE4_dP_48h_HI_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 3-BE4-75-dP-LO-48h_S29,3-BE4-78-dP-LO-48h_S31 \
  -c 3-BE4-75-dP-LIB-48h_S51,3-BE4-78-dP-LIB-48h_S52 \
  -n mageck_counts/3_BE4_dP_48h_LO_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp

mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 37-ABE-B3-dP-HI-48h_S23,37-ABE-F4-dP-HI-48h_S24 \
  -c 37-ABE-B3-dP-LIB-48h_S53,37-ABE-F4-dP-LIB-48h_S54 \
  -n mageck_counts/37_ABE_dP_48h_HI_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 37-ABE-B3-dP-LO-48h_S32,37-ABE-F4-dP-LO-48h_S33 \
  -c 37-ABE-B3-dP-LIB-48h_S53,37-ABE-F4-dP-LIB-48h_S54 \
  -n mageck_counts/37_ABE_dP_48h_LO_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
  
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 37-BE4-A5-dP-HI-48h_S25,37-BE4-B5-dP-HI-48h_S26 \
  -c 37-BE4-A5-dP-LIB-48h_S55,37-BE4-B5-dP-LIB-48h_S56 \
  -n mageck_counts/37_BE4_dP_48h_HI_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
mageck test -k guide_counts/filtered_guideCounts.txt \
  -t 37-BE4-A5-dP-LO-48h_S34,37-BE4-B5-dP-LO-48h_S35 \
  -c 37-BE4-A5-dP-LIB-48h_S55,37-BE4-B5-dP-LIB-48h_S56 \
  -n mageck_counts/37_BE4_dP_48h_LO_vs_LIB \
  --remove-zero both --pdf-report --keep-tmp
```


# save session info
```{r}
sesh.info <- capture.output(sessionInfo())
writeLines(
  sesh.info,
  file.path(project.dir, "session-info", paste0(Sys.Date(), "_BEscreen_240201.txt"))
)
```