---
title: "2_edger_analysis"
author: "Andrea"
date: "26-05-13"
output: html_document
---

# originally adapted from Dane's P300_BE_Screen_edgeR.Rmd script using edger for 
# differential analysis of conditions. This script is then adapted from what was
# done in Nicole's BE screen to compare these together.

# Dane has provided annotations for each BE library which I will use for this
# initial analysis. These are different from what I've done previously in that 
# they are considering all possible annotations TOGETHER, rather than 
# individually. Based on data from PRC2, it is actually likely that these would be 
# occurring simultaeously, so this is a good starting point for analysis before 
# further discussion. Time permitting I can check any guides w/2+ mutations 
# called in CRISPRESSO2 to see the frequency the occur together vs alone and see 
# if there is a higher z-score/FDR for these guides (these are calculated per
# guide and are agnostic to acutal mutation anyway). I could follow up on single
# mutations later on if needs be.

# load libraries
```{r}
library(tidyverse)
library(edgeR)
library(clusterProfiler)
library(RColorBrewer)
library(drawProteins)
library(bartools)
library(viridis)
```

# set dirs
```{r}
project.dir <- "/dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis"
counts.dir <- file.path(project.dir, "guide_counts")
edger.plots.dir <- file.path(project.dir, "plots/2_edger")
edger.dir <- file.path(project.dir, "edger")
ref.dir <- "/dawson_genomics/Projects/P300_scaffolding/BEscreen_250916/reference_files"
setwd(project.dir)
```

# make dir for analysis (always best to make dirs in terminal to ensure proper
# inheritance of permissions)
```{bash}
cd /dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis
mkdir -p plots/2_edger
mkdir edger
```

# import counts files
```{r}
raw.counts <- read.delim(
  file.path(counts.dir, "all_guideCounts.txt"), 
  header = TRUE, 
  sep = "\t", 
  row.names = 1
  )

# reorder cols w/reps & times next to each other
raw.counts <- raw.counts[, c(1, 2, 7, 21, 26, 3, 8, 22, 27, 4, 9, 23, 28, 6, 10, 30, 25, 5, 29, 24, 11, 16, 31, 35, 12, 17, 32, 36, 13, 18, 33, 37, 15, 20, 34, 39, 14, 19, 38)]
head(raw.counts)
```

# generate metadata
```{r}
samp.info <- str_split(colnames(raw.counts)[-1], pattern = "\\.", simplify = T)
metadata <- data.frame(
  "BE" = samp.info[, 2], 
  "Sort" = samp.info[, 5],
  "Time" = gsub("_S[0-9][0-9]", "", samp.info[, 6]),
  "Replicate" = samp.info[, 3],
  "Cells" = samp.info[, 1],
  "Treatment" = samp.info[, 4],
  row.names = colnames(raw.counts)[-1]
  )
metadata$Group <- paste(
  metadata$Cells, 
  metadata$BE, 
  metadata$Treatment, 
  metadata$Sort, 
  metadata$Time, 
  sep = "_"
  )
```

# setup sgRNA info table
```{r}
guides <- fread(
  "/dawson_genomics/Projects/MYC/230926_base-editor_library_analysis/scripts/P300_BE.csv",
  header = FALSE
  )
g.info <- str_split(guides$V1, pattern = "_", simplify = TRUE)
guide.ann <- data.frame(
  "Gene" = g.info[, 1],
  "ID" = guides$V1,
  "Sequence" = guides$V2,
  "Rank" = g.info[, 4]
)
rownames(guide.ann) <- guide.ann$ID
dim(guide.ann)

guide.ann <- guide.ann[rownames(raw.counts), ]
dim(guide.ann)

table(rownames(raw.counts) == rownames(guide.ann))
```

# setup DGEList object
```{r}
dge <- DGEList(
  counts = as.matrix(raw.counts[, -1]), 
  samples = metadata, 
  group = metadata$Group,
  genes = guide.ann, 
  remove.zeros = TRUE
  )
```

# Filter sgRNAs
```{r}
table(dge$samples$group)
# Filter sgRNAs with low counts 
dim(dge)
keep = rowSums(dge$counts > 100) > 5
dge = dge[keep, ]
dim(dge)
```

# reads per sample
```{r}
# Set up group colours, using paired palette but with 3rd for LIB
cols <- dge$samples$group
levels(cols) <- c(
  "#33A02C",
  "#1F78B4",
  "#CAB2D6",
  "#CD6090",
  "#FF6EB4",
  "#E31A1C",
  "#FF8C00",
  "#FFD700",
  "#8B5A2B",
  "#EE9A49",
  "#006400",
  "#27408B",
  "#6A3D9A",
  "#8B3A62",
  "#EE6AA7",
  "#8B0000",
  "#CD6600",
  "#CDAD00",
  "#8B4726",
  "#CD853F"
  ) 
cols <- as.character(cols)

# Plot number of sgRNAs that could be matched per sample 
# and total for each sgRNA across all samples
pdf(
  file.path(edger.plots.dir, "barplots_screen_counts_per_index.pdf"), 
  width = 12, 
  height = 12
  )
par(mar = c(12, 4, 4, 4))
barplot(
  colSums(dge$counts), 
  las = 2, 
  main = "Counts per index", 
  col = cols, 
  cex.names = 0.8,
  cex.axis = 0.8
  )
barplot(
  rowSums(dge$counts), 
  las = 2, 
  main = "Counts per sgRNA",
  cex.names = 0.1, 
  cex.axis = 0.8
  )
dev.off()
```

# MDS plots - all samples
```{r}
# Make an MDS plot to visualise relationships between replicates 
pdf(file.path(edger.plots.dir, "MDS_by_group.pdf"))
plotMDS(dge, labels = dge$samples$Group, col = cols, main = "MDS Plot")
plotMDS(dge, labels = dge$samples$Group, col = cols, main = "MDS Plot", dim.plot = c(1, 3))
plotMDS(dge, labels = dge$samples$Group, col = cols, main = "MDS Plot", dim.plot = c(2, 3))
plotMDS(dge, labels = dge$samples$Group, col = cols, main = "MDS Plot", dim.plot = c(1, 4))
dev.off()
```

# save log2 CPM counts
```{r}
dge.cpm <- edgeR::cpm(dge, log = TRUE)
write.table(
  dge.cpm, 
  file.path(edger.dir, "logCPM_counts_all.txt"), 
  sep = "\t",
  col.names = NA,
  row.names = TRUE,
  quote = FALSE
  )
```

# bartools QC
```{r}
pdf(file.path(edger.plots.dir, "correlation_by_reps.pdf"))
plotBarcodeCorrelation(dge, method = "spearman")
plotBarcodeCorrelation(dge, method = "pearson")
dev.off()
```

# setup DGE model
```{r}
# diff analysis, using GLMs in edgeR as there are more than 2 groups 
# Set up design matrix for GLM
design = model.matrix(~0 + Group, data = dge$samples)
colnames(design) <- gsub("Group", colnames(design), replacement = "")
design
```

# Estimate dispersions
```{r}
xglm = estimateDisp(dge, design, tagwise = TRUE)
sqrt(xglm$common.disp)

pdf(file.path(edger.plots.dir, "BCVplot_all.pdf"))
plotBCV(xglm, main = "BCV Plot")
dev.off()

fit = glmFit(xglm, design)
```

# setup contrasts 
```{r}
contr.mat <- makeContrasts(
  X3_ABE_dP_48h_HIvLIB = X3_ABE_dP_HI_48h - X3_ABE_dP_LIB_48h,
  X37_ABE_dP_48h_HIvLIB = X37_ABE_dP_HI_48h - X37_ABE_dP_LIB_48h,
  X3_ABE_dP_48h_LOvLIB = X3_ABE_dP_LO_48h - X3_ABE_dP_LIB_48h,
  X37_ABE_dP_48h_LOvLIB = X37_ABE_dP_LO_48h - X37_ABE_dP_LIB_48h,
  X3_BE4_dP_48h_HIvLIB = X3_BE4_dP_HI_48h - X3_BE4_dP_LIB_48h,
  X37_BE4_dP_48h_HIvLIB = X37_BE4_dP_HI_48h - X37_BE4_dP_LIB_48h,
  X3_BE4_dP_48h_LOvLIB = X3_BE4_dP_LO_48h - X3_BE4_dP_LIB_48h,
  X37_BE4_dP_48h_LOvLIB = X37_BE4_dP_LO_48h - X37_BE4_dP_LIB_48h,
  ABE_dP_HI_48h_X37vX3 = X37_ABE_dP_HI_48h - X3_ABE_dP_HI_48h,
  ABE_dP_LO_48h_X37vX3 = X37_ABE_dP_LO_48h - X3_ABE_dP_LO_48h,
  BE4_dP_HI_48h_X37vX3 = X37_BE4_dP_HI_48h - X3_BE4_dP_HI_48h,
  BE4_dP_LO_48h_X37vX3 = X37_BE4_dP_LO_48h - X3_BE4_dP_LO_48h,
  levels = design
  )
contr.mat
```

# LRT and smear plot for all 
```{r}
# inititate lists to store tables for each comp
lrt.list <- list()
top4.list <- list()

for(con.i in 1:length(colnames(contr.mat))){
  # Carry out Likelihood ratio test
  lrt.list[[con.i]] = glmLRT(fit, contrast = contr.mat[, con.i])
  # Show top ranked sgRNAs
  topTags(lrt.list[[con.i]], n = 50, sort.by = "logFC")
  top4.list[[con.i]] = topTags(lrt.list[[con.i]], n = Inf)
  # Select P300 sgRNAs with FDR < 0.0001 to highlight on plot
  top4ids = rownames(top4.list[[con.i]]$table[which(top4.list[[con.i]]$table$FDR < 0.01 & top4.list[[con.i]]$table$Gene == "EP300"), ])
  
  # Plot MA (logFC versus logCPM)
  pdf(file.path(edger.plots.dir, paste0("smearPlot_", colnames(contr.mat)[con.i], ".pdf")))
  plotSmear(
    lrt.list[[con.i]], 
    de.tags = top4ids,
    pch = 20, 
    cex = 0.6, 
    ylim = c(-7, 7),
    main = paste(colnames(contr.mat)[con.i], "FDR <0.01 highlighted", sep = "\n")
    )
  abline(h = c(-1, 0, 1), col = c("dodgerblue", "yellow", "dodgerblue"), lty = 2)
  dev.off()
  
  # write out top table
  write.table(
    top4.list[[con.i]]$table,
    file.path(edger.dir, paste0(colnames(contr.mat)[con.i], "_LRT_topTable.txt")),
    sep = "\t",
    quote = FALSE,
    col.names = TRUE,
    row.names = FALSE
  )
}

# add names for referencing
names(lrt.list) <- colnames(contr.mat)
```

# get each BE ann and process as needed
```{r}
# get annotations for both editors
abe.ann <- fread(file.path(ref.dir, "p300.lib.anno.ABE.tsv"))
be4.ann <- fread(file.path(ref.dir, "p300.lib.anno.BE4.tsv"))

# select only columns of interest
abe.ann <- abe.ann[, 1:17]
be4.ann <- be4.ann[, 1:17]

# add columns parsing aa location and substitution into separate columns
abe.prot.loc <- strsplit2(abe.ann$Protein_position, "-")
abe.ann$AA_start <- abe.prot.loc[, 1]
abe.ann$AA_end <- abe.prot.loc[, 2]
abe.aa.change <- strsplit2(abe.ann$Amino_acids, "/")
abe.ann$AA_WT <- abe.aa.change[, 1]
abe.ann$AA2_MUT <- abe.aa.change[, 2]
abe.ann$Editor <- "ABE"
colnames(abe.ann)[c(1, 7)] <- c("Gene", "ENSG")
head(abe.ann)
# same for be4
be4.prot.loc <- strsplit2(be4.ann$Protein_position, "-")
be4.ann$AA_start <- be4.prot.loc[, 1]
be4.ann$AA_end <- be4.prot.loc[, 2]
be4.aa.change <- strsplit2(be4.ann$Amino_acids, "/")
be4.ann$AA_WT <- be4.aa.change[, 1]
be4.ann$AA_MUT <- be4.aa.change[, 2]
be4.ann$Editor <- "BE4"
colnames(be4.ann)[c(1, 7)] <- c("Gene", "ENSG")
head(be4.ann)
```

# incorporate aa. pos and save logFC and Z.score ordered output
```{r}
# inititate list for annotated top hits
top.ann.list <- list()

for(con.i in 1:length(colnames(contr.mat))){
  # calc Z score for P300 guides
  top.ann.list[[con.i]] <- as.data.frame(top4.list[[con.i]]) %>% dplyr::filter(Gene == "EP300")
  top.ann.list[[con.i]]$Z.score <- (top.ann.list[[con.i]]$logFC - mean(top.ann.list[[con.i]]$logFC))/sd(top.ann.list[[con.i]]$logFC)
  dim(top.ann.list[[con.i]]) 
  # summary stats on Z score distribution
  print(summary(top.ann.list[[con.i]]$Z.score))
  plot(density(top.ann.list[[con.i]]$Z.score), main = colnames(contr.mat)[con.i])
  plot(density(top.ann.list[[con.i]]$logFC), main = colnames(contr.mat)[con.i])
  # join aa pos info into P300 table
  top.ann.list[[con.i]]$Rank <- as.numeric(top.ann.list[[con.i]]$Rank)
  # join w/ann for BE used
  comp.info <- str_split(colnames(contr.mat)[con.i], pattern = "_", simplify = T)
  ifelse(
    comp.info[, 2] == "ABE" | comp.info[, 1] == "ABE", 
    top.ann.list[[con.i]] <- left_join(top.ann.list[[con.i]], abe.ann, by = c("Gene", "ID", "Sequence", "Rank")),
    top.ann.list[[con.i]] <- left_join(top.ann.list[[con.i]], be4.ann, by = c("Gene", "ID", "Sequence", "Rank"))
    )


  # write P300 aa table 
  write.table(
    top.ann.list[[con.i]],
    file.path(edger.dir, paste0(colnames(contr.mat)[con.i], "_P300_top_Zscore_MUTann.txt")),
    sep = "\t",
    quote = FALSE,
    col.names = TRUE,
    row.names = FALSE
  )
}
names(top.ann.list) <- colnames(contr.mat)
```


# plot Z score, logCPM and FDR across protein body by BE
```{r}
for(con.i in 1:length(colnames(contr.mat))){
  # need to cast start pos as numeric
  top.ann.list[[con.i]]$AA_start <- as.numeric(top.ann.list[[con.i]]$AA_start)
  # z-score
  pdf(file.path(edger.plots.dir, paste0("P300_pos_Zscore_", colnames(contr.mat)[con.i], ".pdf")))
  top.ann.list[[con.i]] %>% 
    ggplot(aes(x = AA_start, y = Z.score, color = FDR)) + 
    geom_point() + 
    geom_hline(yintercept = c(-4, 4), color = "red3") + 
    scale_color_continuous(low = "blue4", high = "grey90") +
    ggtitle(colnames(contr.mat)[con.i]) +
    scale_x_continuous(breaks = seq(0, 2414, by = 200)) +
    theme_light()
  dev.off()
  # logCPM
  pdf(file.path(edger.plots.dir, paste0("P300_pos_logFC_", colnames(contr.mat)[con.i], ".pdf")))
  top.ann.list[[con.i]] %>% 
    ggplot(aes(x = AA_start, y = logFC, color = FDR)) + 
    geom_point() + 
    scale_color_continuous(low = "blue4", high = "grey90") +
    ggtitle(colnames(contr.mat)[con.i]) +
    theme_light()
  dev.off()
  # FDR
  pdf(file.path(edger.plots.dir, paste0("P300_pos_FDR_", colnames(contr.mat)[con.i], ".pdf")))
  top.ann.list[[con.i]] %>% 
    ggplot(aes(x = AA_start, y = FDR, color = FDR)) + 
    geom_point() + 
    geom_hline(yintercept = 0.05, color = "red3") + 
    scale_color_continuous(low = "blue4", high = "grey90") +
    ggtitle(colnames(contr.mat)[con.i]) +
    theme_light()
  dev.off()
}
```

# draw P300 structure using drawProteins package
```{r}
# UniProt accession numbers of P300 = Q09472
p300.json <- drawProteins::get_features("Q09472")
p300.ann <- drawProteins::feature_to_dataframe(p300.json)
head(p300.ann)
```

# plot domains alongside screen data
```{r}
# subset rel data
p300.ann %>% dplyr::filter(type == "CHAIN")
unique(p300.ann$type)
#prot.data <- p300.ann %>% filter(type == "DOMAIN" | type == "REGION" & description != "Disordered")

# region & domain plot
prot.plot <- draw_canvas(p300.ann)
prot.plot <- draw_chains(prot.plot, p300.ann, label_chains = F) + 
  coord_cartesian(xlim = c(0, 2414))
prot.plot <- draw_regions(prot.plot, p300.ann, show.legend = TRUE) + 
  theme(legend.position = "bottom")
prot.plot <- draw_domains(prot.plot, p300.ann, label_size = 2) + 
  theme_classic() + 
  theme(legend.position = "bottom", 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.line.y = element_blank()) 
```

# wrangle variant annotation and make colour scheme to include these in plot
##### From Nicole's analysis, similar for this experiment #####
# To simplify variants I only took the first in the list (after looking through
# all unique categories in all comparisons they appear to be prioritised by
# severity, although missense_variant > splice_region_variant??). Example list 
# prior to wrangling (all ABE had 17 & BE4 had 16 unique combinations):
#unique(top.ann.list[[1]]$Consequence)
 [1] "missense_variant"                                                          
 [2] "None"                                                                      
 [3] "splice_acceptor_variant"                                                   
 [4] "splice_region_variant,coding_sequence_variant"                             
 [5] "splice_donor_variant,coding_sequence_variant"                              
 [6] "splice_donor_region_variant,intron_variant"                                
 [7] "synonymous_variant"                                                        
 [8] "splice_donor_5th_base_variant,intron_variant"                              
 [9] "splice_acceptor_variant,splice_polypyrimidine_tract_variant,intron_variant"
[10] "missense_variant,splice_region_variant"                                    
[11] "intron_variant"                                                            
[12] "splice_region_variant,synonymous_variant"                                  
[13] "splice_polypyrimidine_tract_variant,intron_variant"                        
[14] "splice_region_variant,intron_variant"                                      
[15] "splice_region_variant,splice_polypyrimidine_tract_variant,intron_variant"  
[16] "start_lost,5_prime_UTR_variant"                                            
[17] "5_prime_UTR_variant"  
# Then for any splicing variants I changed to all just be categorized
# as "splice_variant". 

# setting colour scheme for these categories.
```{r}
consequences <- c(
  "start_lost",
  "missense_variant",
  "splice_variant",
  "5_prime_UTR_variant",
  "intron_variant",
  "synonymous_variant",
  "None"
)
# Dark2 color palette except for replaced gray with blue
cons_color <-
  list(
    "#8db600",
    "#be0032",
    "#f38400",
    "#f3c300",
    "#875692",
    "#a1caf1",
    "#848482"
  )
names(cons_color) <- consequences
```

# need to wrangle each consequence column while iterating through the list items
# then plot z-scores w/ variant colouring and protein annotation
```{r}
for(con.i in 1:length(colnames(contr.mat))){
  # split consequence and add single consequence to df
  cons.sing <- strsplit2(top.ann.list[[con.i]]$Consequence, ",")
  top.ann.list[[con.i]]$Consequence_single <- cons.sing[, 1]
  # change any splice change to the same
  splice.var <- which(substr(top.ann.list[[con.i]]$Consequence_single, 1, 7) == "splice_")
  top.ann.list[[con.i]]$Consequence_single[splice.var] <- "splice_variant"
  # need to cast AA_start as numeric
  top.ann.list[[con.i]]$AA_start <- as.numeric(top.ann.list[[con.i]]$AA_start)
  
  # make combined plot plot
  screen.plot <- top.ann.list[[con.i]] %>% 
    ggplot(aes(x = AA_start, y = Z.score, color = Consequence_single, label = ID)) + 
    geom_point() + 
    geom_text_repel(data = subset(top.ann.list[[con.i]], abs(Z.score) > 4)) +
    geom_hline(yintercept = c(-4, 4), color = "red3") + 
    scale_color_manual(values = cons_color) +
    ggtitle(colnames(contr.mat)[con.i]) +
    scale_x_continuous(breaks = seq(0, 2414, by = 200)) +
    theme_light()

  plt <- list(screen.plot = screen.plot, prot.plot = prot.plot)
  ggpubr::ggarrange(
    plt$screen.plot, 
    plt$prot.plot, 
    heights = c(12, 4),
    ncol = 1, 
    nrow = 2, 
    align = "v", 
    legend = "bottom", 
    common.legend = TRUE
    )
  
  ggsave(file.path(
    edger.plots.dir, 
    paste0("P300_prot_Zscore_", colnames(contr.mat)[con.i], ".pdf")
    ), width = 8, height = 5)
}
```

# plot stats for control guides as well for comparison
```{r}
# make violins of LFC and FDR for each "Gene"
for(con.i in 1:ncol(contr.mat)){
  pdf(file.path(edger.plots.dir, paste0("violin_FDR_", colnames(contr.mat)[con.i], "_edger.pdf")))
  top4.list[[con.i]] %>% 
    ggplot(aes(x = Gene, y = FDR, fill = Gene)) + 
    geom_violin() + 
    geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
    geom_hline(yintercept = 0.05, color = "firebrick4") + 
    scale_fill_viridis(discrete = TRUE) +
    ggtitle(colnames(contr.mat)[con.i]) +
    theme_light() 
  dev.off()
  
  pdf(file.path(edger.plots.dir, paste0("violin_LFC_", colnames(contr.mat)[con.i], "_edger.pdf")))
  top4.list[[con.i]] %>% 
    ggplot(aes(x = Gene, y = logFC, fill = Gene)) + 
    geom_violin() + 
    geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
    scale_fill_viridis(discrete = TRUE) +
    ggtitle(colnames(contr.mat)[con.i]) +
    theme_light() 
  dev.off()
}
```

# save R object
```{r}
save.image(file.path(project.dir, "2_edger_analysis.RData"))
```


# save session info
```{r}
sesh.info <- capture.output(sessionInfo())
writeLines(
  sesh.info,
  file.path(project.dir, "session-info", paste0(Sys.Date(), "_edger_analysis_250916.txt"))
)
```