---
title: "3_mageck_analysis"
author: "Andrea"
date: "2026-05-21"
output: html_document
---

# previously in Ollie's BE screen edger and mageck were quite similar, but I 
# would preference mageck as the underlying assumptions are more suitable. 
# However, in Nicole's screen mageck had loads of FDR significant hits, whereas
# edger only had a few so it's better to use the edger results and avoid false
# positives. Still running mageck on Dane's data here for comparison. 
# INPUT: RData object from previous script (2_edger_analysis.RData)

project dir: /dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis

# load libraries and set paths
```{r}
library(data.table)
library(viridis)

project.dir <- "/dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis"
mageck.dir <- file.path(project.dir, "mageck_counts")
mag.ann.dir <- file.path(project.dir, "mageck_results_ann")
mageck.plots.dir <- file.path(project.dir, "plots/3_mageck")
setwd(project.dir)
```

# load data from previous script output containing annotations
```{r}
load(file.path(project.dir, "2_edger_analysis.RData"))
```

# make dir for ann files & plots
```{bash}
cd /dawson_genomics/Projects/MYC/240201_P300_BE_reactivation_screen/Andrea_analysis
mkdir mageck_results_ann
mkdir plots/3_mageck
```

# annotate mageck counts and write out
```{r}
# annotate each set of mageck results
mag.files <- list.files(mageck.dir, pattern = "sgrna_summary.txt", full.names = TRUE)
# initiate variables to save tables
mag.list <- list()
top.list <- list()
for(mag.i in 1:length(mag.files)){
  # get mageck results and merge w/ann
  mag.list[[mag.i]] <- fread(mag.files[mag.i])
  mag.list[[mag.i]] <- merge(mag.list[[mag.i]], guide.ann, by.x = c("Gene", "sgrna"), by.y = c("Gene", "ID"))
  # cast rank as numeric for joining
  mag.list[[mag.i]]$Rank <- as.numeric(mag.list[[mag.i]]$Rank)
  # join w/ann for BE used
  comp.info <- str_split(basename(mag.files[mag.i]), pattern = "_", simplify = T)
  ifelse(
    comp.info[, 2] == "ABE", 
    mag.list[[mag.i]] <- merge(
      mag.list[[mag.i]], 
      abe.ann, 
      by.x = c("Gene", "sgrna", "Sequence", "Rank"), 
      by.y = c("Gene", "ID", "Sequence", "Rank")
      ),
    mag.list[[mag.i]] <- merge(
      mag.list[[mag.i]],
      be4.ann, 
      by.x = c("Gene", "sgrna", "Sequence", "Rank"), 
      by.y = c("Gene", "ID", "Sequence", "Rank")
      )
    )
  # calculate z-score of LFC
  mag.list[[mag.i]]$Z_score <- (mag.list[[mag.i]]$LFC - mean(mag.list[[mag.i]]$LFC))/sd(mag.list[[mag.i]]$LFC)
  # save full ann table
  write.table(
    mag.list[[mag.i]],
    file.path(mag.ann.dir, gsub(".sgrna_summary.txt", "_all_mageck_ann.txt", basename(mag.files[[mag.i]]))),
    sep = "\t",
    col.names = TRUE,
    row.names = FALSE,
    quote = FALSE
  )
  # get only those in P300 >=2 sds either way 
  top.list[[mag.i]] <- subset(mag.list[[mag.i]], Gene == "EP300" & abs(Z_score) >= 2)
  # check how many in each
  print(dim(top.list[[mag.i]]))
  # write out top table
  write.table(
    top.list[[mag.i]],
    file.path(mag.ann.dir, gsub(".sgrna_summary.txt", "_P300_Zscore2_mageck_ann.txt", basename(mag.files[[mag.i]]))),
    sep = "\t",
    col.names = TRUE,
    row.names = FALSE,
    quote = FALSE
  )
}
names(mag.list) <- gsub(".sgrna_summary.txt", "", basename(mag.files))
```

# make z-score protein plots for mageck stats w/cons colouring as in edger plots
```{r}
# protein plot already defined in 2_edger_analysis.Rmd
for(mag.i in 1:length(mag.files)){
  # split consequence and add single consequence to df
  cons.sing <- strsplit2(mag.list[[mag.i]]$Consequence, ",")
  mag.list[[mag.i]]$Consequence_single <- cons.sing[, 1]
  # change any splice change to the same
  splice.var <- which(substr(mag.list[[mag.i]]$Consequence_single, 1, 7) == "splice_")
  mag.list[[mag.i]]$Consequence_single[splice.var] <- "splice_variant"

  # need to cast start pos as numeric
  mag.list[[mag.i]]$AA_start <- as.numeric(mag.list[[mag.i]]$AA_start)
  # screen plot
  screen.plot <- mag.list[[mag.i]] %>% dplyr::filter(Gene == "EP300") %>% 
      ggplot(aes(x = AA_start, y = Z_score, color = Consequence_single, label = sgrna)) + 
      geom_point() + 
      geom_text_repel(data = subset(mag.list[[mag.i]], abs(Z_score) > 4), max.overlaps = Inf) +
      geom_hline(yintercept = c(-4, 4), color = "red3") + 
      scale_color_manual(values = cons_color) +
      ggtitle(names(mag.list)[mag.i]) +
      theme_light() + 
      theme(legend.position = "right") + 
      coord_cartesian(xlim = c(0, 2414)) + 
      theme(axis.title.x = element_blank())

  plt <- list(screen.plot = screen.plot, prot.plot = prot.plot)
  ggpubr::ggarrange(
    plt$screen.plot, 
    plt$prot.plot,
    heights = c(12, 4),
    ncol = 1, 
    nrow = 2, 
    align = "v", 
    legend = "bottom", 
    common.legend = TRUE
    )
  
  ggsave(
    file.path(
      mageck.plots.dir, 
      paste0("P300_prot_Zscore_", names(mag.list)[mag.i], "_mageck.pdf")
      ), 
    width = 8, 
    height = 5
    )
}
```

# also make FDR plots from mageck data
```{r}
for(mag.i in 1:length(mag.files)){
  pdf(file.path(mageck.plots.dir, paste0("P300_pos_FDR_", names(mag.list)[mag.i], "_mageck.pdf")))
  mag.list[[mag.i]] %>% 
    ggplot(aes(x = AA_start, y = FDR, color = FDR)) + 
    geom_point() + 
    geom_hline(yintercept = 0.05, color = "red3") + 
    scale_color_continuous(low = "blue4", high = "grey90") +
    ggtitle(names(mag.list)[mag.i]) +
    theme_light()
  dev.off()
}
```

# also plot control guides for comparison
```{r}
# make violins of LFC and FDR for each "Gene"
for(mag.i in 1:length(mag.files)){
  pdf(file.path(mageck.plots.dir, paste0("violin_FDR_", names(mag.list)[mag.i], "_mageck.pdf")))
  mag.list[[mag.i]] %>% 
    ggplot(aes(x = Gene, y = FDR, fill = Gene)) + 
    geom_violin() + 
    geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
    geom_hline(yintercept = 0.05, color = "firebrick4") + 
    scale_fill_viridis(discrete = TRUE) +
    ggtitle(names(mag.list)[mag.i]) +
    theme_light() 
  dev.off()
  
  pdf(file.path(mageck.plots.dir, paste0("violin_LFC_", names(mag.list)[mag.i], "_mageck.pdf")))
  mag.list[[mag.i]] %>% 
    ggplot(aes(x = Gene, y = LFC, fill = Gene)) + 
    geom_violin() + 
    geom_boxplot(width = 0.1, color = "grey", alpha = 0.2) +
    scale_fill_viridis(discrete = TRUE) +
    ggtitle(names(mag.list)[mag.i]) +
    theme_light() 
  dev.off()
}
```

# make a scatterplot w/corr for HIvLO LFC
```{r}
ct <- cor.test(mag.list$`3_ABE_dP_48h_HI_vs_LIB`$LFC, mag.list$`3_ABE_dP_48h_LO_vs_LIB`$LFC)
linear_model <- lm(mag.list$`3_ABE_dP_48h_HI_vs_LIB`$LFC ~ mag.list$`3_ABE_dP_48h_LO_vs_LIB`$LFC)
pdf(file.path(mageck.plots.dir, "Scatterplot_3_ABE_dP_48hr_LFC.pdf"))
plot(
  mag.list$`3_ABE_dP_48h_HI_vs_LIB`$LFC,
  mag.list$`3_ABE_dP_48h_LO_vs_LIB`$LFC, 
  col = rgb(0.4, 0.4, 0.8, 0.6, alpha = 0.5),
  xlab = "3_ABE_dP_48hr_HI",
  ylab = "3_ABE_dP_48hr_LO"
  )
abline(linear_model, col = "red")
text(0, 1.5, paste0("R = ", round(ct$estimate, 4)))
text(0, 1, paste0("p <= ", round(ct$p.value, 5)))
dev.off()

ct <- cor.test(mag.list$`3_BE4_dP_48h_HI_vs_LIB`$LFC, mag.list$`3_BE4_dP_48h_LO_vs_LIB`$LFC)
linear_model <- lm(mag.list$`3_BE4_dP_48h_HI_vs_LIB`$LFC ~ mag.list$`3_BE4_dP_48h_LO_vs_LIB`$LFC)
pdf(file.path(mageck.plots.dir, "Scatterplot_3_BE4_dP_48hr_LFC.pdf"))
plot(
  mag.list$`3_BE4_dP_48h_HI_vs_LIB`$LFC,
  mag.list$`3_BE4_dP_48h_LO_vs_LIB`$LFC, 
  col = rgb(0.4, 0.4, 0.8, 0.6, alpha = 0.5),
  xlab = "3_BE4_dP_48hr_HI",
  ylab = "3_BE4_dP_48h_LO"
  )
abline(linear_model, col = "red")
text(0, 1.2, paste0("R = ", round(ct$estimate, 4)))
text(0, 1, paste0("p <= ", round(ct$p.value, 5)))
dev.off()

ct <- cor.test(mag.list$`37_ABE_dP_48h_HI_vs_LIB`$LFC, mag.list$`37_ABE_dP_48h_LO_vs_LIB`$LFC)
linear_model <- lm(mag.list$`37_ABE_dP_48h_HI_vs_LIB`$LFC ~ mag.list$`37_ABE_dP_48h_LO_vs_LIB`$LFC)
pdf(file.path(mageck.plots.dir, "Scatterplot_37_ABE_dP_48hr_LFC.pdf"))
plot(
  mag.list$`37_ABE_dP_48h_HI_vs_LIB`$LFC,
  mag.list$`37_ABE_dP_48h_LO_vs_LIB`$LFC, 
  col = rgb(0.4, 0.4, 0.8, 0.6, alpha = 0.5),
  xlab = "37_ABE_dP_48h_HI",
  ylab = "37_ABE_dP_48h_LO"
  )
abline(linear_model, col = "red")
text(0, 1.75, paste0("R = ", round(ct$estimate, 4)))
text(0, 1.5, paste0("p <= ", round(ct$p.value, 5)))
dev.off()

ct <- cor.test(mag.list$`37_BE4_dP_48h_HI_vs_LIB`$LFC, mag.list$`37_BE4_dP_48h_LO_vs_LIB`$LFC)
linear_model <- lm(mag.list$`37_BE4_dP_48h_HI_vs_LIB`$LFC ~ mag.list$`37_BE4_dP_48h_LO_vs_LIB`$LFC)
pdf(file.path(mageck.plots.dir, "Scatterplot_37_BE4_dP_48hr_LFC.pdf"))
plot(
  mag.list$`37_BE4_dP_48h_HI_vs_LIB`$LFC,
  mag.list$`37_BE4_dP_48h_LO_vs_LIB`$LFC, 
  col = rgb(0.4, 0.4, 0.8, 0.6, alpha = 0.5),
  xlab = "37_BE4_dP_48h_HI",
  ylab = "37_BE4_dP_48h_LO"
  )
abline(linear_model, col = "red")
text(0, 1.2, paste0("R = ", round(ct$estimate, 4)))
text(0, 1, paste0("p <= ", round(ct$p.value, 5)))
dev.off()
```

# save R object
```{r}
save.image(file.path(project.dir, "3_mageck_analysis.RData"))
```

# save session info
```{r}
sesh.info <- capture.output(sessionInfo())
writeLines(
  sesh.info,
  file.path(project.dir, "session-info", paste0(Sys.Date(), "_mageck_analysis.txt"))
)
```