---
title: "ChIPseq_240424"
author: "Andrea"
date: "2024/05/13"
output: html_document
---

fastq location: /pipeline/Runs/NextSeq/240424_VH01624_92_AAFJTHNM5/ProjectFolders/Project_Jesse-Balic/
project directory: /dawson_genomics/Projects/MYC/240424_ChIP/

# set up project dir and process (fastqc, trim, align, tdfs)
```{bash}
mkdir alignment
mkdir fastqc
mkdir trimmed_fastqs
mkdir logs
mkdir scripts
mkdir tdfs
mkdir fastq_screen

for fq in /pipeline/Runs/NextSeq/240424_VH01624_92_AAFJTHNM5/ProjectFolders/Project_Jesse-Balic/Sample*/*_R1_001.fastq.gz
do
sbatch scripts/alignHg38BDGP6.sbatch $fq
done
```

# run fastq screen to check spike-in proportion
```{bash}
for fq in /pipeline/Runs/NextSeq/240424_VH01624_92_AAFJTHNM5/ProjectFolders/Project_Jesse-Balic/Sample*/*_R1_001.fastq.gz
do
sbatch scripts/fastq_screen.sbatch $fq 
done

module load multiqc/1.8
multiqc -ip .
# make multiqc w/o Dm files
multiqc -ip -n noDM_multiqc --ignore-samples *Dm* .
```

# peak calling (both narrow and broad) include bedgraph output, use input from 240307_ChIP
```{bash}
mkdir macs/narrowpeaks
mkdir macs/broadpeaks

sbatch scripts/macs2.sbatch alignment/H3K27ac-DMSO-R2-RSI_S85.sort.rmdup.bam ../240307_ChIP/alignment/Input-DMSO-R1_S146.sort.rmdup.bam
sbatch scripts/macs2.sbatch alignment/H3K27ac-A485-R2-RSI_S86.sort.rmdup.bam ../240307_ChIP/alignment/Input-A485-R1_S147.sort.rmdup.bam
sbatch scripts/macs2.sbatch alignment/H3K27ac-dTAG-R2-RSI_S87.sort.rmdup.bam ../240307_ChIP/alignment/Input-dTAG-R1_S148.sort.rmdup.bam
sbatch scripts/macs2.sbatch alignment/H3K27ac-A485-dTAG-R2-RSI_S88.sort.rmdup.bam ../240307_ChIP/alignment/Input-A485-dTAG-R1_S150.sort.rmdup.bam
```

# convert MACS bedgraphs to bigwigs as in C&R
```{bash}
mkdir bigwigs

for bdg in macs/broadpeaks/*_treat_pileup.bdg
do
sbatch scripts/bedgraphToBigwig.sbatch $bdg
done
```

# make genome wide plot (MANE) and 6 lists heatmap w/o spike-in
```{bash}
mkdir results
mkdir plots

sbatch scripts/DTMatrix_H3K27ac_6lists_MACSnorm.sbatch
sbatch scripts/DTMatrix_H3K27ac_MANE_MACSnorm.sbatch
```

# get bam stats for spike in 
```{bash}
module load samtools/1.9

for bam in alignment/*_Dm.sort.rmdup.bam
do
echo $bam >> alignment/DmBamFlagstat.txt
samtools flagstat $bam >> alignment/DmBamFlagstat.txt
done
```

# Load R libraries and set paths
```{r}
library(data.table)

project.dir <- "/dawson_genomics/Projects/MYC/240424_ChIP"
align.dir <- file.path(project.dir, "alignment")
peaks.dir <- file.path(project.dir, "macs/broadpeaks")
```

# collate DM counts, use lib sizes to get scale factor for bigwigs
```{r}
dm.stat <- read.table(file.path(align.dir, "DmBamFlagstat.txt"), header = FALSE, fill = TRUE)
dm.files <- dm.stat[grep("Dm.sort.rmdup.bam",dm.stat$V1), 1]
dm.reads <- dm.stat[grep("total", t(dm.stat$V5)), 1]

# make text file with bam names and scale factors to use for deeptools
# scale factor for each is 10000000/Dm reads
bam.sf <- cbind(gsub("_Dm.sort.rmdup.bam", ".sort.rmdup.bam", dm.files), 10000000/as.numeric(dm.reads))
write.table(
  bam.sf,
  file.path(align.dir, "bam_sf.txt"),
  sep = ":",
  col.names = FALSE,
  row.names = FALSE,
  quote = FALSE
)
```

# make bigwigs norm by spike-in
```{bash}
while read -r line
do
IFS=: read -r bam sf <<< $line
sbatch scripts/DeeptoolsBamCov.sbatch $bam $sf
done < alignment/bam_sf.txt
```

# make deeptools plots for H2BK16ac using spike-in bigwigs
```{bash}
sbatch scripts/DTMatrix_H3K27ac_MANE_Dmnorm.sbatch
sbatch scripts/DTMatrix_H3K27ac_6lists_Dmnorm.sbatch
```

# also use spike-in to make reactivated vs non plots
```{bash}
sbatch scripts/DTMatrix_H3K27ac_2genelists_Dmnorm.sbatch
```

# make 6 lists heatmap sorted by P300 combo
```{bash}
sbatch scripts/DTMatrix_H3K27ac_6lists_Dmnorm_P300comboSort.sbatch
```

# redo MANE and 6lists w/ new MANE annotation
```{bash}
sbatch scripts/DTMatrix_H3K27ac_MANE_Dmnorm.sbatch
sbatch scripts/DTMatrix_H3K27ac_6lists_Dmnorm_P300comboSort.sbatch
```

# save session info
```{r}
sesh.info <- capture.output(sessionInfo())
writeLines(
  sesh.info,
  file.path(project.dir, "session-info", paste0(Sys.Date(), "_ChIP_240424.txt"))
)
```
